A low-mortality benchmark for the United States

### 2016-03-06 BPLE/Manuscript.tex
started working in new cod fractions

\title{A best practices composite lifetable for the United States}

@@

\begin{abstract}
We calculate a preliminary series of best-possible lifetables for the United
States from 1959 to 2004, defined on the basis of the age-specific aggregate of
the lowest observed age-cause specific death rates among the 50
states and the District of Columbia. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
average 1.9 and 2.2 years higher than the highest state life expectancy in each year for males and females, respectively. We argue that the US states best practices lifetable is a useful
gauge of future mortality trends and a good benchmark against which to compare
state mortality patterns.
\ackn{The work reported in this manuscript was supported in part by the U.S.
National Institute On Aging of the National Institutes of Health under award
numbers R01-AG011552 and R01-AG040245. The content is solely the responsibility of the authors and does not necessarily represent the official views of the funding agencies.}
\end{abstract}

@@

The level and timing of cause-specific responses to particular
technological or well-being improvements varies between populations. Therefore,
the lowest observed all-cause force of mortality is not necessarily composed of
the lowest cause-specific forces of mortality, which means that the vanguard life
expectancy may not reflect the best mortality possible given current conditions
in a set of populations. For instance, the lowest rate of suicide and heart
disease at age 45 will not necessarily be found in the same population. As an
alternative to vanguard life expectancy, one could create cause-specific
low-mortality benchmarks, or more synthetically, combine the lowest mortality
observed by age and cause into a hybrid minimum mortality schedule. This synthetic low mortality lifetable--- hybrid with respect to reference population over age and cause--- was already suggested by \citet{wunsch1975minimum} and \citet{vallin2008minimum}. These authors investigated trends at the national-level. We refer to such hybrid minimum
lifetables as ``best practices'' lifetables. \citet{eikemo2014} use the term
``best practices'' in a similar sense of an ideal mix of conditions, but do so
via combining particular risk factors with their estimated mortality impacts to
produce best case scenarios. The end effect is quite similar, to the extent that
either estimation strategy shows what society could achieve given perfect
diffusion of some definition of best conditions. In the case of a best practices
lifetable, we condition on mortality outcomes only, making the endeavor much
simpler in practice.

National populations are heterogeneous with respect to many factors, both
observed and unobserved, that affect cause-specific mortality rates. This does
not make international mortality comparisons futile, but it may make perfect
convergence unimaginable even in the very long run. This is the case for both
all cause and cause-specific mortality. When projecting mortality for a given country, we wonder whether mortality from an external population, or set of populations, is the best ``sanity
check'' for mortality forecasts.
For this reason, we propose a minimum mortality trend for the United States
calculated on the basis of its own territorial subdivisions, the 50 states plus
the District of Columbia. This choice limits the sources of unaccounted
heterogeneity in international comparisons, and it also guarantees uniformity of
data sources.
To be clear, we do not necessarily propose using a trend in best practices mortality as a component of coherent mortality forecasts, and we do
not undertake such an exercise. Instead, we view the U.S. best practices
mortality as indicative of the potential already present in the U.S., a special
kind of ``nowcast''.

\section*{Data and methods}
All death count data come from the public NCHS mortality microdata files
\citep{NCHSdata}.\footnote{The current analysis is based on the public files
with geographic information only up to 2004. We have bene granted access to
geographic variables in a NCHS Research Data Center, and will extend analysis
through 2013 for the final paper version.} Population denominators come from a
beta version of the U.S.
States project of the \citet{HMD}. Further processing of
population counts into exposures is done according to the HMD Methods
Protocol, and all-cause lifetables were calculated by single ages \citep{HMDMP}.
Death counts for causes were aggregated into quinquennial age groups by year and sex for eleven large groups that minimize disruptions between ICD transitions, and then
converted to fractions. Fractions were then multiplied into the single-age
all-cause death rates. The eleven causes of death considered in this abstract
version are displayed in Table~\ref{tab:desc}, with their relative percentages.

% latex table generated in R 3.1.2 by xtable 1.7-4 package
% Wed Sep 23 17:46:55 2015
\begin{table}[ht]
\centering
\caption{Eleven causes of death considered in the current study version,
percentages.}
\label{tab:desc}
\begin{tabular}{rrr}
 & Females & Males \\ 
  \hline
All other & 29.91 & 33.95 \\ 
  Breast & 3.73 & 0.03 \\ 
  Cardiovascular & 38.49 & 38.10 \\ 
  Cerebrovascular & 10.50 & 6.84 \\ 
  Lung & 3.55 & 6.50 \\ 
  Other Cerebrovascular & 0.33 & 0.28 \\ 
  Other Malignant Neoplasms & 11.01 & 9.83 \\ 
  Other smoking & 0.57 & 1.36 \\ 
  Prostate &  & 2.25 \\ 
  Stomach & 0.67 & 0.87 \\ 
  Uterus & 1.25 &  \\ 
   \hline
\end{tabular}
\end{table}

>>>>

\title{A low-mortality benchmark for the United States}

@@

\begin{abstract}
We calculate a preliminary series of best-possible lifetables for the United
States from 1959 to 2004, defined on the basis of the age-specific aggregate of
the lowest observed age-cause specific death rates among the 50
states and the District of Columbia. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
average 1.9 and 2.2 years higher than the highest state life expectancy in each
year for males and females, respectively. We argue that the US states
low-morality lifetable is a useful gauge of future mortality trends. We apply
the low-mortality as a benchmark to compare state mortality
patterns and quantify the potential for immediately acheivable improvement.
\ackn{The work reported in this manuscript was supported in part by the U.S.
National Institute On Aging of the National Institutes of Health under award
numbers R01-AG011552 and R01-AG040245. The content is solely the responsibility of the authors and does not necessarily represent the official views of the funding agencies.}
\end{abstract}

@@

The lowest observed all-cause force of mortality is not
necessarily composed of the lowest cause-specific forces of mortality. This
means that the vanguard life expectancy may not reflect the best mortality possible given current conditions in a set of populations. For instance, the lowest rate of suicide and heart
disease at age 45 will not necessarily be found in the same population. As an
alternative to vanguard life expectancy, one could create cause-specific
low-mortality benchmarks, or more synthetically, combine the lowest mortality
observed by age and cause into a hybrid minimum mortality schedule. This synthetic low mortality lifetable--- hybrid with respect to reference population over age and cause--- was already suggested by \citet{wunsch1975minimum} and \citet{vallin2008minimum}. These authors investigated trends at the national-level. We refer to such hybrid minimum
lifetables as ``best practices'' lifetables. \citet{eikemo2014} use the term
``best practices'' in a similar sense of an ideal mix of conditions, but do so
via combining particular risk factors with their estimated mortality impacts to
produce best case scenarios. The end effect is quite similar, to the extent that
either estimation strategy shows what society could achieve given perfect
diffusion of some definition of best conditions. In the case of a best practices
lifetable, we condition on mortality outcomes only, making the endeavor much
simpler in practice.

National populations are heterogeneous with respect to many factors, both
observed and unobserved, that affect cause-specific mortality rates. The level and timing of cause-specific responses to particular
technological or well-being improvements varies between populations. This does
not make international mortality comparisons futile, but it may make perfect
convergence unimaginable even in the very long run. This is the case for both
all cause and cause-specific mortality. When projecting mortality for a given country, we wonder whether mortality from an external population, or set of populations, is the best ``sanity
check'' for mortality forecasts.
For this reason, we propose a minimum mortality trend for the United States
calculated on the basis of its own territorial subdivisions, the 50 states plus
the District of Columbia. This choice limits the sources of unaccounted
heterogeneity in international comparisons, and it also guarantees uniformity of
data sources.
To be clear, we do not necessarily propose using a trend in best practices mortality as a component of coherent mortality forecasts, and we do
not undertake such an exercise. Instead, we view the U.S. best practices
mortality as indicative of the potential already present in the U.S., a special
kind of ``nowcast''.

\section*{Data and methods}
All death count data come from the restricted NCHS mortality microdata files
\citep{NCHSdata}.\footnote{We will extend analysis
through 2013 for the final paper version.} Single-age population denominators
and complete state lifetables come from a beta version of the U.S.
States project of the \citet{HMD}. Death counts by causes of death are first
grouped into 30 large cause groups and then converted to fractions in abridged
age groups, $[0,1)$, $[1,4)$, $[5,9)$, \ldots, $100+$ \citep{HMDMP}. Fractions
in abridged age groups are then multiplied into single-age all-cause death
rates to produce a first estimate of single-age cause-specific death rates.
The 30 causes of death considered in this abstract
version are displayed in Table~\ref{tab:desc}, with their relative percentages.

% latex table generated in R 3.1.2 by xtable 1.7-4 package
% Wed Sep 23 17:46:55 2015
%\begin{table}[ht]
%\centering
%\caption{Eleven causes of death considered in the current study version,
%percentages.}
%\label{tab:desc}
%\begin{tabular}{rrr}
% & Females & Males \\ 
%  \hline
%All other & 29.91 & 33.95 \\ 
%  Breast & 3.73 & 0.03 \\ 
%  Cardiovascular & 38.49 & 38.10 \\ 
%  Cerebrovascular & 10.50 & 6.84 \\ 
%  Lung & 3.55 & 6.50 \\ 
%  Other Cerebrovascular & 0.33 & 0.28 \\ 
%  Other Malignant Neoplasms & 11.01 & 9.83 \\ 
%  Other smoking & 0.57 & 1.36 \\ 
%  Prostate &  & 2.25 \\ 
%  Stomach & 0.67 & 0.87 \\ 
%  Uterus & 1.25 &  \\ 
 %  \hline
%\end{tabular}
%\end{table}
